---
title: "covid_19_analysis"
output: pdf_document
date: '2023-05-02'
---

Setup the data from github and import the datasets.
```{r setup}
library(tidyverse)
library(readr)
library(knitr)
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",  
                "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)

us_cases <- read_csv(urls[1])
us_deaths <- read_csv(urls[2])
```

Clean the data to appropriate form by dropping irrelevant columns and change the date data type. I also needed to pivot the datasets to long form so that it can be used for analysis.
```{r tidy_data}
us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths <- us_deaths %>%
    pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

# Join the two datasets together
us <- us_cases %>%
  full_join(us_deaths)
```

Prepare datasets for further visualizations.
```{r viz}
# create dataset that include total cases and deaths for each state
us_by_state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases,
         deaths, deaths_per_mill, Population) %>%
  ungroup()

# dataset that include total us cases and deaths
us_total <- us_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date, cases,
         deaths, deaths_per_mill, Population) %>%
  ungroup()

# visualize the amount of Covid19 cases and deaths in US overtime
us_viz <- us_total %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45)) +
  labs(title = "Covid19 in US")
us_viz
```

Use a linear regression model to evaluate the difference on deaths_per_mill, between Florida and Colorado overtime.
```{r model}
flo_colo <- us_by_state %>% 
  filter(Province_State %in% c("Florida", "Colorado"))

flo_colo$states <- as.factor(flo_colo$Province_State)

# fit linear regression model
lr_model <- lm(deaths_per_mill ~ states * date, data = flo_colo)

# set predicted value
flo_colo$predicted_deaths_per_mill <- predict(lr_model)

ggplot(flo_colo, aes(x = date, y = deaths_per_mill, color = states)) +
  geom_line(size = 1) +
  geom_line(aes(y = predicted_deaths_per_mill), size = 1, linetype = "dashed") +
  labs(title = "Covid19 death per mill compared by states",
       x = "Date",
       y = "Deaths per million") +
  theme_bw()

summary(lr_model)
```

We can see that state is statistically significant predictive variable for death_per_mill. Interestingly, Florida shows a higher death rate as predicted comparing to Colorado, what could be the reason for it? Could it be the public health policies, population density, or temperature? 


